Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-18 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
First experiments description reviewed

  This is the invariant part of the framework, but other than that,
  the algorithm can be run in many different ways: 
  stopping when a solution is found or, as we have done in this paper,
  using Web Workers. In fact, since it is a pool-based system such as
  the one described in \cite{LNCS86720702}, any kind of client that
  calls the application programming interface (API) can be used, % (Paloma) I thought that a client implements an API, or uses an API, or even works with an API. But never heard of a client which "works" an API. Is this correct?
  written in any kind of language. However, in this paper we are going
  to focus on the dynamics and measurements of the volunteer part of
  the system.

@@

The researcher only has to change a function, which can be written in
JavaScript or in any other language with cross-compilation to
JavaScript \cite{web:compilersjs} to solve a different
problem. In this case the classical Trap function \cite{Ackley1987} has been
used as well as the Rastrigin's floating-point optimization
problem. JavaScript is a functional language and declaring a different 
function and handing at the creation of the algorithm object, called
{\tt Classic}, is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):
\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in Node.js and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a non-blocking single thread
  allows the service of many requests. In fact, a limit in the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability. 
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system, or even browser requirement as long
  as Javascript is enabled: anyone
  visiting the page, even from mobile devices, can load the algorithm.
  In case the browser does not support HTML5 Web workers (e.g., 
  Windows Phone or some version of Android) or their use
  has been disabled, a basic version of {\sf NodIO} can also be used.
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through the autonomous
  operation of every individual island without any synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server. This architecture 
  has been tested in other high demand systems with success, but 
  in the case of {\sf NodIO}, additional experiments are needed to asses 
  the scalability of the communication system. % (Paloma) Is it ok if we talk about scalability when we were talking about adaptiveness? Scalability was also described in the first point.
  %(Mario) Scalability of the communication system in particular?
\item Since the algorithm runs in the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources than the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which makes the operation totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL, available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is better than what you would obtain in
  your own desktop system. If this is not the case, it is a pointless
  academic exercise.
\end{itemize}

@@

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of Internet Protocol addressess (IPs)
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. % (Paloma) Feature or disadvantage?
 However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in red). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, and of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it drops sharply implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}; we obviously do
not pretend to compare our system in scale or complexity with it, only
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which appears also in our system. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, negative towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next.

>>>>

  This is the invariant part of the framework, but other than that,
  the algorithm can be run in many different ways: 
  stopping when a solution is found or, as we have done in this paper,
  using Web Workers. In fact, since it is a pool-based system such as
  the one described in \cite{LNCS86720702}, any kind of client that
  works the application programming interface (API) can be used, % (Paloma) I thought that a client implements an API, or uses an API, or even works with an API. But never heard of a client which "works" an API. Is this correct?
  written in any kind of language. However, in this paper we are going
  to focus on the dynamics and measurements of the volunteer part of
  the system.

@@

The researcher only has to change a function, which can be written in
JavaScript or in any other language with cross-compilation to
JavaScript \cite{web:compilersjs} to solve a different
problem. In this case the classical Trap function \cite{Ackley1987} has been
used as well as the Rastrigin's floating-point optimization
problem. JavaScript is a functional language and declaring a different 
function and handing at the creation of the algorithm object, called
{\tt Classic}, is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):
\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in Node.js and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a non-blocking single thread
  allows the service of many requests. In fact, a limit in the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability. 
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system, or even browser requirement as long
  as Javascript is enabled: anyone
  visiting the page, even from mobile devices, can load the algorithm.
  In case the browser does not support HTML5 Web workers (e.g., 
  Windows Phone or some version of Android) or their use
  has been disabled, a basic version of {\sf NodIO} can also be used.
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through the autonomous
  operation of every individual island without any synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server. This architecture 
  has been tested in other high demand systems with success, but 
  in the case of {\sf NodIO}, additional experiments are needed to asses 
  the scalability of the system. % (Paloma) Is it ok if we talk about scalability when we were talking about adaptiveness? Scalability was also described in the first point.
\item Since the algorithm runs in the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources than the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which makes the operation totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL, available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is better than what you would obtain in
  your own desktop system. If this is not the case, it is a pointless
  academic exercise.
\end{itemize}

@@

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of Internet Protocol addressess (IPs)
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP {\tt PUT}s per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. % (Paloma) Feature or disadvantage?
 However, it puts a probably
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of {\tt PUT}s per unique IP and fit to a Generalized
  Extreme Value distribution (in red). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows a S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, of the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in the
distribution of HTTP {\tt PUT}s, equivalent to the number of
generations divided by 100, contributed by every user, which is shown
in Figure \ref{fig:puts:os}. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table*}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of {\tt PUT}s per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our system in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our system. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
into account the low correlation between successive events shown
above.

At any rate, this also shows that a convenient way of increasing the
computing power would be increasing this minimum amount per user. This
issue led us to create the next version of the framework, which is
presented next. % (Paloma) This sentences are sort of alone here... Could we extend a bit the first "this" to connect them to the previous paragraph ?
